Estimating demographic bias on tests of children's early vocabulary

### 2022-02-01 paper/cogsci.Rmd
tweaking

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to accurately measure early language. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
    measures of children's aggregate early language skill. However, CDIs contain
    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
    often by children of varying backgrounds. This study used a database of American English 
    CDIs to identify words that showed strong bias for particular demographic groups of children, on 
    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
    eliminating these items reduced the expected ability difference between groups. Additionally,
    we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
    sex-based word learning bias, and discuss possible sources of demographic bias in early word learning.

@@

Researchers, clinicians, and parents have long been fascinated with the surprising speed and variability in the growth of young children's vocabulary.
Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and individual differences in the aptitude of the child -- including potential language deficits.
Children show both consistency in some skills across development, as well as significant influence from external factors.
For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
Yet maternal education, often used as a proxy for socioeconomic status (SES), has also been found to be associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and to be predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for a review].
Other demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@frank2021; @eriksson2012differences] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].

However, it is also difficult to get a complete measure of young children's language skill: long recordings are prohibitively difficult to collect and transcribe [see the singular exception: @roy2015predicting], and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

We focus on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

@@

While many languages show similar demographic effects on early word learning [e.g., the female advantage: @eriksson2012differences; and the first-born advantage: @frank2021], the size of these demographic effects vary across languages [see @frank2021, Ch. 6]. 
Why might we see such variation?
While it is possible that there is an underlying difference in mean language ability between demographic groups, and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the tests are biased to some degree--and perhaps to a different degree in different languages?
The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
The validity of any test with many items favoring one group over another (say, children from rural vs. urban households) is questionable, as the test may be overestimating the ability of test-takers in the former (rural) group, and/or underestimating the ability of the latter group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
Of course, there may in fact be some meaningful ability difference between demographic groups, as is likely the case for the female language advntage, which has been documented in many languages [@eriksson2012differences; @frank2021].
However, these demographic differences in early language learning have not been investigated as potentially due to DIF.
That is the contribution of the present study.
Specifically, our goal is to test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs. female), maternal education (no more than secondary vs. at least some college), and race (white vs. non-white).

@@

where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.

@@

## Demographic effects in baseline Rasch model
A baseline Rasch model was fitted to the entire dataset, without demographic information.
Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects. 
Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$).
High-SES children had higher language ability than low-SES children ($\beta=0.23$, $p=.02$), an advantage that grew with age ($\beta=0.11$, $p<.001$).
White children had higher language ability than non-white children ($\beta=0.50$, $p<.001$), an advantage that grew with age ($\beta=0.08$, $p<.001$).
We will re-examine these demographic regressions after trimming items showing extreme DIF.

@@

```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs. age by demographic group, from the baseline Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors) +
  theme(legend.position = "top")

p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") +
  scale_color_discrete(name = "SES", type=ses_colors) + 
  theme(legend.position = "top")

p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(name = "ethnicity", type=eth_colors) + 
  theme(legend.position = "top")

ggarrange(p1, p2, p3, ncol=3, nrow=1)
```

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

>>>>

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to measure early language accurately. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide reliable and valid
    measures of children's aggregate early language skill. However, CDIs contain
    hundreds of vocabulary items, some of which may not be heard (and thus learned) equally
    often by children of varying backgrounds. This study used a database of American English 
    CDIs to identify words that showed strong bias for particular demographic groups of children, on 
    dimensions of sex (male vs. female), race (white vs. non-white), and maternal education 
    (high vs. low). For each dimension, we identified dozens of strongly biased items, and showed that 
    eliminating these items reduced the expected ability difference between groups. Additionally,
    we investigated how well the relative frequency of words spoken to young girls vs. boys predicted 
    sex-based word learning bias, and discuss possible sources of demographic differences in early word learning.

@@

Researchers, clinicians, and parents have long been fascinated with the surprising speed and variability in the growth of young children's vocabulary.
Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and individual differences in the aptitude of the child -- including potential language deficits.
Children show both consistency in some skills across development, as well as significant influence from external factors.
For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.

Yet there are consistent demographic predictors of early vocabulary. Maternal education, often used as a proxy for socioeconomic status (SES), is associated with children's language processing and vocabulary by 18 months [@fernald2013ses], and is predictive of later educational outcomes [@marchman2008speed; see @schwab2016 for review].
Other factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@frank2021; @eriksson2012differences] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].

One challenge in comprehensive assessment of these differences is that is difficult to get a complete measure of young children's language skill. Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

We focus here on the MacArthur-Bates Communicative Development Inventories [CDIs; @fenson1994; @Fenson2007], a set of parent-reported measures of children's productive and receptive language skills, which offer a low-cost and reliable way to estimate children's early language skills [@fenson1994].
CDIs have shown good predictive validity [e.g., @fenson1994; @bornstein2012stability; @duff2015infant].
Our primary focus is the vocabulary checklist portion of the CDI Words & Sentences (CDI:WS) form, comprised of 680 early-learned words across 22 categories (e.g., animals, vehicles, action words, pronouns) selected to assess the productive vocabulary of children 16 to 30 months of age.
For each item on the CDI:WS, caregivers are asked to respond whether the target child has been heard to say (i.e. produce) the given item.
Children's total vocabulary score on the CDI:WS is tightly correlated with other facets of early language (e.g., grammatical competence and gesture), suggesting that the language system is "tightly woven" [@frank2021].
Due to these desirable properties, CDIs have been adapted to dozens of languages, and a central repository of CDI data contributed from all over the world has been created [Wordbank; @frank2017; @frank2021].

@@

@@

@@

```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "t", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs. age by demographic group, from the baseline Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors) +
  theme(legend.position = "top")

p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") +
  scale_color_discrete(name = "Maternal Ed.", type=ses_colors) + 
  theme(legend.position = "top")

p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(name = "Race", type=eth_colors) + 
  theme(legend.position = "top")

ggarrange(p1, p2, p3, ncol=3, nrow=1)
```
CDIs are a useful tool for measuring demographic effects because of their wide use across languages. 
<!-- While many languages show similar demographic effects on early word learning [e.g., the female advantage: @eriksson2012differences; and the first-born advantage: @frank2021], the size of these demographic effects vary across languages [see @frank2021, Ch. 6].  -->
<!-- Why might we see such variation? -->
But their use in this context poses a chicken and egg problem. 
While it is possible that there is an underlying difference in mean language ability between demographic groups, CDIs could also have bias for or against particular groups based on the words they include. 
<!-- If a hypothetical vocabulary test only consisted of words for cold-weather gear, it would naturally show bias against children who grew up in hot climates (or who were born at certain times of year).  -->
<!-- , and that cross-linguistic variation is driven by differences in the language environments of different cultures, could it also be possible that the  -->

The idea that some items on a test may show bias, favoring one group over another, is known in psychometrics as Differential Item Function (DIF; @holland1993differential]. 
DIF can decrease the validity of a test, as the test may be overestimating the ability of test-takers in one group and underestimating the ability of another group -- not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
For example, if a vocabulary test is composed almost entirely of the names of farm equipment, the test will underestimate the knowledge of urban children, who have experience of other contexts, despite being less familiar with farm equipment than their rural peers.
Of course, there may be an ability difference between demographic groups, as is likely the case for the female language advantage, which has been documented in many languages [@eriksson2012differences; @frank2021]. And both observations could be true -- a test could be biased to inflate a true demographic difference.

Our goal in the current study is to use DIF analyses to assess these possibilities. 
Specifically, we test the 680 vocabulary items on the American CDI:WS for DIF along three dichotomized demographic dimensions: sex (male vs. female), maternal education (no more than secondary vs. at least some college), and race (white vs. non-white).

\noindent where $D$ is a constant scaling parameter ($D=1.702$) which makes the logistic closely match the ogive function in traditional factor analysis [@R-mirt; @reckase2009].
Child ability ($\theta$) and item difficulty ($b$) distributions are standardized (i.e., mean of 0), and expected to be normally-distributed. 
Children with high latent ability ($\theta$) will be more likely to produce any given item than children with lower latent ability, and more difficult items will be produced by fewer children (at any given $\theta$) than easier items.

## Demographic effects in baseline Rasch model
We fit a baseline Rasch model to the entire dataset, without demographic information.
Figure 1 shows children's language ability vs. age by demographic group from the baseline Rasch model, which assumes no DIF (i.e., equal item parameters for all groups).
A linear regression for each demographic group, with age (centered) and its interaction, showed significant effects. 
Female children had higher language ability than male children ($\beta=0.56$, $p<.001$), with no significant interaction with age ($\beta=0.02$, $p=.10$).
High-SES children had higher language ability than low-SES children ($\beta=0.23$, $p=.02$), an advantage that grew with age ($\beta=0.11$, $p<.001$).
White children had higher language ability than non-white children ($\beta=0.50$, $p<.001$), an advantage that grew with age ($\beta=0.08$, $p<.001$).
We will re-examine these demographic regressions after trimming items showing extreme DIF.

@@

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.
